---
title: "Lab 01 Python setup"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: false
---

***

# 1. Install `Anaconda`

We will use a distribution of Python called `Anaconda`, which includes a range of useful packages for scientific coding, such as `matplotlib`, `numpy`, and `pandas.` It all comes with the `conda` package manager - a tool for easily installing other Python add-on packages that you may want to use. It also comes with a few useful programs which can be used to write Python code.

Download [`Anaconda`](https://www.anaconda.com/products/individual). Make sure to install a **Python 3.x** version specific to your operating system.

***

# 2. `python` via `Spyder`

**1.** Once you installed `Anaconda`, open `Spyder`.

**2.** In the Console window, type: 

```{r, eval=F}
print("Hello World!")
```

A `Hello World!` output message will show up. 

**3.** Now use `python` as a calculator. In the Console window, type the following lines one by one: 

```{r, eval=F}
1+2
1-2
1*2
1/2
1+2*3
(1+2)*3
```

**4.** Now type the following line in the Console window: 
```{r, eval=F}
1==2
```

What is your result? Can you explain why?
