---
title: "Lab 04"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: false
---

***

# 1. Daily new cases of COVID-19  

In this lab, we will look at the daily number of newly reported cases of COVID-19 worldwide, compiled by at the [European Centre for Disease Prevention and Control (ECDC)](https://www.ecdc.europa.eu/en/publications-data/download-todays-data-geographic-distribution-covid-19-cases-worldwide). We will fit an ARIMA model to the time series, and make predictions for the coming days. 

Through this demo, you will learn the necessary steps to fitting ARIMA models to time series data:

* plotting the data

* possibly transforming the data

* identifying the dependence orders of the model

* parameter estimation

* making predictions

***

**1.1 Load the libraries** 

Load `dplyr`, `lubridate`, and `forecast` packages.

**1.2 Load the daily new cases data** 

Download the `.csv` data from [here](https://zhu-group.github.io/ese5023/download/data_20201108.csv), take a quick glance at the file. 

Then move the file (`data_20201108.csv`， ~ 4MB) to your `working directory`, load the file using `read.csv()`, and convert it to a `tibble` (name it as `COVID_tbl`) using the `as_tibble()` function. 

**1.3 Get the daily total number of newly reported cases worldwide ** 

What you just read from `data_20201108.csv` is the daily number of new cases of COVID-19 by country worldwide (the `cases` column). Think about how to get daily **total** number of new cases of COVID-19 over the world.

[Hint] You may find the following lines useful:

```{r, eval=F}
...
  mutate(dateRep = as.Date(dateRep, format = '%d/%m/%Y')) %>% 
  group_by(dateRep) %>% 
...
```

**1.4 Plot the data**

Plot the daily new cases, what do you observe? Do you see an exponential trend? Why is that?

**1.5 Filter the data**

As you can see from the above figure, daily new case starts to increase exponentially since late March and early April. To make the time series more reliable (a.k.a, not effected by initial fluctuations), let's only use the data after April 1st, 2020. Do this using the `filter()` function.

**1.6 Convert a vector to a time series**

Convert the vector from **1.5** to a time series using the `ts()` function. As we are dealing with daily data here, use *Julian Day* as the indicator of the time series. You can return the Julian Day of a certain date using the `yday()` function from the `lubridate` package. 

Copy the following lines to your script, make sure you understand them before running.

```{r, eval=F}
# Start date of the time series, read from the .csv file
Date_start <- as.Date("2020-04-01")
# End date of the time series, read from the .csv file
Date_end   <- as.Date("2020-11-08")
# Get the Julian Day of the start date
JD_start   <- yday(Date_start)
# Convert the vector data to a time series 
global_cases_ts <- ts(COVID_tbl$global_cases, 
                      start=c(2020,JD_start),
                      frequency=365)
# The indicator of the time series
inds            <- seq(Date_start, Date_end, by = "day")
```

Plot the time series, what do you observe? 

**1.7 Transform the time series**

Take `log` to the `global_cases_ts` time series, and assign the results to a new object `global_cases_ts_log`. Plot `global_cases_ts_log`, also check acf and pacf of `global_cases_ts_log`. Do you see a trend in the time series `global_cases_ts_log`? And is this time series stationary?

**1.8 Take the difference**

One way to make a non-stationary time series stationary is to take the *difference*. For an ARIMA(`p`,`d`,`q`) model, `d=1` if difference is taken once, and `d=2` if difference is taken twice. In R, this is done with the `diff()` function:

```{r, eval=F}
global_cases_ts_log_d1 <- diff(global_cases_ts_log)
plot(global_cases_ts_log_d1)
```

What do you see from the time series plot? Does the time series look stationary now?

**1.9 Auto ARIMA fitting**

In the `forecast` package, R provides a very useful function called `auto.arima()`, which returns the best ARIMA model according to either AIC, AICc or BIC value. The function conducts a search over possible models within the order constraints provided. 

Run the following lines:
```{r, eval=F}
# Automated forecasting using an ARIMA model
model <- auto.arima(global_cases_ts_log)
model
```

What is the best ARIMA model?

**1.10 Make predictions**

Once you get an ARIMA model, you can use it to predict values, new cases in this demo, in the future time steps.

Run the following lines:
```{r, eval=F}
# Number of days to predict
days_forecast  <- 15
# Number of include in the plot
days_in_plot   <- 30
# Make predictions using the forecast() function
forecast_15days <- forecast(model, days_forecast)
# Plot
plot(forecast(model, days_forecast), include=days_in_plot, 
     xlab="Time", ylab="log(global cases)",type="o",lwd=2) 
```

Here the blue line is the estimated values, the blue shaded area shows the 80% confidence interval, and the gray area shows the 95% confidence interval.

Change `days_forecast` to, see how the predictions would vary.

**1.11 Get predicted values**

What is the predicted value on Nov. 09? And What is the predicted value on Nov. 30? Report estimated values and their 80% confidence intervals.

**1.12 Verify the predictions**

Now download the latest data from [ECDC](https://opendata.ecdc.europa.eu/covid19/casedistribution/csv/data.csv), compute the global total number of new cases of COVID-19 on Nov. 09. Compare this number with the predicted value from **1.11**. Do you get a reasonable prediction?

**1.13 [optional] Even more: Fit a SARIMA model**

In fact, the `global_cases_ts_log` shows weekly cycles, this is not well considered in the model. You can make an even better ARIMA model by removing this pattern, or by fitting a SARIMA model, where "S" stands for seasonal. Look at [Package ‘sarima’](https://cran.r-project.org/web/packages/sarima/sarima.pdf) for more.
