---
title: "Lab 01"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: false
---

***

# 1. Merge branches with GitKraken

In this exercise, we will learn how to create a branch, modify it, and merge it to the `main` with GitKraken.

**1.** Create a local repository named `Lab1_Exercise` with GitKraken.

**2.** Modify the `README` file by adding one sentence, stage and then commit the change.

**3.** Add one text file named `temperature.txt`, stage and then commit the change.

**4.** Modify `temperature.txt` by adding the following line: `The air temperature is 29 degree today`. Stage and then commit the change.

**5.** Right-click the graph node of the latest commit (green circle on the middle panel), select `Create branch here`, 
then enter the branch name `branch1`.

**6.** When you create a new branch, GitKraken will automatically switch to the branch. You can always check which branch you are at by looking at the `branch` drop-down menu at the top left corner. Something like `main` or `branch1` (you just create it!). You can switch between different branches simply by selecting the one you want to work on. Such an action is called `checkout` in Git. For now, let's stay on `branch1`.

**7.** Add one text file named `aqi.txt`, stage and then commit the change.

**8.** Modify `aqi.txt` by adding the following line: `The air quality index is 53 today`. Stage and then commit the change.

**9.** Modify the `README` file by adding the following line: `I made this change on branch1`, stage and then commit the change.

**10.** By far, we have added one file (`aqi.txt`) and modified one file (`README`), we think it's good enough to go and we want to incorporate those changes in the `main` branch. This step is called `merge` in Git. Now switch to `main` by selecting it from the the `branch` drop-down menu at the top left corner. You will see `branch1` is `3` commits ahead of `main`, this is because we only worked on `branch1` but not `main`. Right-click `branch1` tab on the middle panel, select `Merge branch1 into main`. At this time, we have successfully merged the changes made on `branch1` into `main`. 

**11.** Modify the `README` file again by adding the following line: `I made this change on main`, stage and then commit the change. 

**12.** Right-click the graph node of the latest commit (green circle on the middle panel), select `Create branch here`, 
then enter the branch name `branch2`. You may find that GitKraken automatically switchs to the `branch2`.

**13.** Modify `temperature.txt`, change the line to: `The air temperature is 32 degree today`. Stage and then commit the change.

**14.** Now switch back to `main` by selecting it from the the `branch` drop-down menu at the top left corner. Open the `temperature.txt` file in your local repo, what is the temperature read form your file? `32` or `29`? Can you explain why?

**15.** Modify `temperature.txt`, change the line to: `The air temperature is 35 degree today`. Stage and then commit the change.

**16.** So far, we modified the same line in `temperature.txt` differently on two branches (`35` on `main` while `32` on `branch2`). As you can imagine, when we merge two differently modified files, a conflict will rise. Right-click `branch2` tab on the middle panel, select `Merge branch2 into main`, GitKraken will pop out an error message `A file conflict was found when attemping to merge into main`. 

**17.** Click `temperature.txt` on the `Conflicted Files`, choose `The air temperature is 32 degree today` as the output, then save, and click `Commit and Merge`. By doing so, we have successfully resolved the conflicts.

***

# 2. Merge branches with Github

In this exercise, we will learn how to create a branch, modify it, and merge it to the `main` on GitHub.

**1.** Go through [Understanding the GitHub flow](https://guides.github.com/introduction/flow/).

**2.** Open GitKraken, push `Lab1_Exercise` to GitHub. Click `1 branche` next the `main` drop-down menu for a quick overview about the branches.

**3.** Now follow [GitHub guide](https://guides.github.com/activities/hello-world/) step `2` to create a new branch `branch3`.

**4.** Follow step `3` to modify `README`  by adding the following line: `I made this change on branch3 via GitHub`, commit the change.

**5.** Follow step `4` to open a Pull Request.

**6.** Follow step `5` to merge your Pull Request.

**7.** Open GitKraken and pull `Lab1_Exercise` from GitHub. Check the flow. 

**8.** Finally, go back to `Lab1_Exercise` on GitHub, click `Settings`, then scorll down to the bottom, click `Delete this repository`.

***

# 3. Install `Anaconda`

We will use a distribution of Python called `Anaconda`, which includes a range of useful packages for scientific coding, such as `matplotlib`, `numpy`, and `pandas.` It all comes with the `conda` package manager - a tool for easily installing other Python add-on packages that you may want to use. It also comes with a few useful programmes which can be used to write Python code.

Download [`Anaconda`](https://www.anaconda.com/products/individual). Make sure to install a **Python 3.X** version specific to your operating system.

***

# 4. `python` via `Spyder`

**1.** Once you installed `Anaconda`, open `Spyder`.

**2.** In the Console window, type: 

```{r, eval=F}
print("Hello World!")
```

A `Hello World!` output message will show up. 

**3.** Now use `python` as a calculator. In the Console window, type the following lines one by one: 

```{r, eval=F}
1+2
1-2
1*2
1/2
1+2*3
(1+2)*3
```

**4.** Now type the following line in the Console window: 
```{r, eval=F}
1==2
```

What is your result? Can you explain why?
