---
title: "Lab 01"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: false
---

***

# 1. Permutations and Combinations 

* Install the `gtools` package

* Use `?` to learn how to use `combinations()` and `permutations()`

* Use `permutations()` to show all possible permutations of picking two balls with replacement/repetition from a bag with a `red`, `blue`, `green`, and `black` ball

* Use `combinations()` to show all possible combinations of picking two balls with replacement/repetition from a bag with a `red`, `blue`, `green`, and `black` ball

***

# 2.More about Mauna Loa CO~2~

* Use `Keeling_Data` in [Section 03](https://zhu-group.github.io/ese5023/Section_03.html#Loading_csv_file), compute the annual mean of CO~2~ since 1959, plot your results.

* The annual mean rate of growth of CO~2~ in a given year is defined as the difference in concentration between the end of December and the start of January of that year. [^1] Compute the annual mean rate of growth  of CO~2~ since 1959. Plot the annual mean rate of growth. What do you observe? Does the increasing trend accelerate or decelerate?

[^1]: [Trends in Atmospheric Carbon Dioxide](https://www.esrl.noaa.gov/gmd/ccgg/trends/gr.html)

***

# 3. Return a value from a function

```{r, eval=FALSE}
# Define the function FindTop()
FindTop        <- function(N){
    MoviePage  <- readLines('https://movie.douban.com/top250')
    Pattarn    <- '<span class=\"title\">&nbsp;/&nbsp;([^<]*)</span>'
    Lines      <- grep(Pattarn, MoviePage, value=TRUE)
    Lines2     <- sub(Pattarn,  '\\1', Lines )
    Names      <- sub('^ *', '', Lines2)
    print(Names[1:N])
}
# Call the function with arguement 5 and 10
FindTop(5)
FindTop(10)
```
* Run the above lines. 

* Go to [this website](https://movie.douban.com/top250), figure out what the above function does. 

* For now, don't worry about the weird syntax in getting `Pattarn`, `Lines2`, and `Names`. Make sure you understand how to use `readLines()`, `grep()`, and `sub()` by reading their help documents. 

* What's the difference between `read.csv()` and `readLines()`?

Many a times, we will require our functions to do some processing and **return back the result**. This is accomplished with the `return()` function in R. 

Change the above function to:
```{r}
# Define the function FindTop_new()
FindTop_new    <- function(N){
    MoviePage  <- readLines('https://movie.douban.com/top250')
    Pattarn    <- '<span class=\"title\">&nbsp;/&nbsp;([^<]*)</span>'
    Lines      <- grep(Pattarn, MoviePage, value=TRUE)
    Lines2     <- sub(Pattarn,  '\\1', Lines )
    Names      <- sub('^ *', '', Lines2)
    return(Names[1:N])
}
```
* Now you should be able to pass the output of `FindTop_new()` to a vector, say `MoiveList`. Try this. 

# 4. Git platform and client

* Sign up for [GitHub](https://github.com): choose a user name that isn’t already taken, provide an email address and a password, and click the big green “Sign up for GitHub” button.

* Install [GitKraken](https://www.gitkraken.com/git-client)
