---
title: "Lab 02"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: false
---

***

# 1. The gapminder data

In this exercise, we look at some of the data on wealth and life expectancy of countries overtime used by Hans Rosling in his famous [TED talk](https://www.ted.com/talks/hans_rosling_the_best_stats_you_ve_ever_seen). 

The data has been conveniently packaged in the R package `gapminder`, a subset of the original data set available at [Gapminder](https://www.gapminder.org/). For each of 142 countries, the package provides values for life expectancy, GDP per capita, and population, every five years, from 1952 to 2007. [^1]

[^1]: [ggplot2 tutorial: gapminder data](http://euclid.psych.yorku.ca/www/psy6135/tutorials/gapminder.html)

**1.1** Install the `gapminder` package if you haven't. Then load the data by:
```{r, eval=FALSE}
# Load the gapminder library
library(gapminder)
```
Using the `class()` function to verify `gapminder` is a `tibble`.

You are required to use functions from `tidyr`, `dplyr`, and `ggplot2` to finish the following questions:

**1.2** Plot life expectancy of China from 1952 to 2007.

**1.3** Plot population of countries in Africa from 1952 to 2007, show each country in a separate panel (facet). Based on your observations, which country has experienced the fastest growth in population?

**1.4** Calculate the average GDP per capita from 1952 to 2007 for each country. Which country has the highest average GDP per capita? And which one has the lowest?

**1.5** Calculate the standard error of the life expectancy for each continent.

**1.6** Plot the difference in total GDP (defined as `population * GDP per capita`) between the USA and China from 1952 to 2007.

**1.7** Create a `tibble` contains only `country`, `year`, and `pop`, named it as `Country_Pop`. Create another `tibble` named `Country_LifeExp` that contains `country`, `year`, and `lifeExp`. Run:
```{r, eval=FALSE}
# Save objects to a RData file
save(Country_Pop, Country_LifeExp, file="My.RData")
```
Check your `working direction`, where should now have a new file `My.RData`. 

Here we have used the `save()` function to save two R objects to a specified file (`My.RData` in this case). Using `?` to learn more about how to use the `save()` function. 

**1.8** Close RStudio. Re-open it, and run:
```{r, eval=FALSE}
# Load a RData file
load("My.RData")
```
Here we call the `load()` function to read the two R objects (`Country_Pop` and `Country_LifeExp`) back from the file. Use the `ls()` function to check. Consider saving your temporary data as `.RData` files to save time.

**1.9** Run:
```{r, eval=FALSE}
# Save objects to a .csv file
write.csv(Country_LifeExp, file="Country_LifeExp.csv")
```
Check your `working direction`, you should now find a new file `Country_LifeExp.csv`. Check this file. Using `?` to learn more about how to use the `write.csv()` function. 

**1.10** Now save your results from 1.4 to a `.csv` file using the `write.csv()` function. 

***

# References  